***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**

### 2015-10-19 Schloss_Cluster_PeerJ_2015.Rmd
Insert citations

---
title: "***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**"
author: Patrick D. Schloss* and Sarah L. Westcott
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

## To Do:
* references
* revise
* add sensitivity/specificity of closed-reference searching

@@

* To whom correspondence should be addressed.
pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI

\newpage
## Abstract

@@

The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples. To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins. These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@schloss2011assessing]. Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g. @human2012structure], soil [e.g. @shade2013streptomycin], aquatic ecosystems [e.g. @gilbert2012defining], and numerous other environments. Within the field of microbial ecology, a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%. One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project. However, with the dissemination of clustering methods within software such as mothur [@schloss2009introducing], QIIME [@Caporaso2010], and other tools [@edgar2010search, @sun2009esprit, @10.7717/peerj.593], it is important to understand how different clustering methods implement this conventional OTU threshold. Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs. Broadly speaking, three approaches have been developed to assign sequences to OTUs.

The first approach has been referred to as phylotyping [@schloss2011assessing] or closed reference clustering [@navas2013advancing]. This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@schloss2011assessing]. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.

The second approach has been referred to as distance-based [@schloss2011assessing] or *de novo* clustering [@navas2013advancing]. In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures then these problems can be overcome [@Kozich2013]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@edgar2010search, @sun2009esprit, @10.7717/peerj.593]. One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [REFS]. Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011]. For this reason, *de novo* clustering has been preferred across the field.

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@Navas2013]. Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions employed by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset (REFS). These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering. Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data (REFS). This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists. Furthermore, it is well known that the rates of evolution across lineages are not the same (REFS). We recently proposed an approach for evaluating OTU assignments using the distance between sequences (REF). Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient (MCC; REF). This is an objective approach to assessing the quality of the OTU assignments.

A recent analysis by He and colleagues (REF) attempted to characterize the three general clustering approaches by focusing on what they called stability. They defined stability as the ability of an method to provide the same clustering on a subset of the data as was found in the full dataset. Their concept of stability did not account for the accuracy of the OTU assignments and instead focused on the precision of the assignments. A method may be very precise, but low in accuracy. In the current analysis, we assessed the accuracy and precision of the various clustering methods. Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of accuracy. In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.

@@

***Summary and replication of He study.*** We obtained the Canadian soil dataset from Roesch et al. (REF) and processed the sequences as described by He and colleagues. Using these data, we reconsidered three of the more critical analyses performed in the He study.

@@

Third, the authors attempted to describe the effects of the OTU assignment instability on comparisons of communities. They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset when only using the unstable OTUs (REFS). Although they were able to detect significant p-values, they appeared to be of marginal biological significance. Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different. The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL). Regardless of the statistical or biological significance of these results, the analysis does not make sense since, by definition, representing communities based on their unstable OTUs would yield differences. Furthermore, the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds. To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets. The justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable. It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels. Finally, the authors built off of this analysis to count the number of OTUs that were differentially represented between the subsetted and full datasets by each method. This analysis assumes that the OTUs generated using the full dataset were correct, which was an unsubstantiated assumption since the authors did not assess the quality of the OTU assignments. Because this analysis was so poorly designed, we did not seek to reproduce it.

***OTU assignment methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance bewteen the sequences. To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth. This enabled us to calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C). The SL and CL methods were the most sensitive; however, the accuracy of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods. Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D). Third, with the possible exception of the CL method, the MCC values for each of the only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same method, they produce OTU assignments that are equally good. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the the full dataset is not optimally assigned to OTUs. Our analysis indicates that the best method for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL method.

@@

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods. First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permit the sequencing of millions of sequences for a study. In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS). The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods. Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene (REFS). With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces (REFS). The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% (REFS). Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points. First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the He dataset (Figure 3AB). With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`. Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the Canadian soil dataset for all of the methods except the CL and SL methods. Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD). This result suggests that the optimal method is likely to be database-dependent. Finally, as was observed with the Canadian soil dataset, there was little variation in MCC values observed among the 30 randomizations. Therefore, although the methods have a stochastic component, the OTU assignments do not vary meaningfully between runs. The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the assignments are highly reproducible. Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods are consistently more robust than either of the reference-based methods.

***Evalution of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods. As originally described in the He study, the AGC and DGC methods utilized the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to generate candidate alignments whereas VSEARCH generates the actual global alignments. The VSEARCH developers claim that this difference enhances the sensitivity of VSEARCH relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC methods, as implemented by the two programs, yielded OTU assignments of similar quality. In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the methods (Figure 4). When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-verion of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets. In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable. These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC methods.

@@

Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference methods to generate OTU labels that are used in several downstream applications. It is commonly used to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses (REFS) and to identify reference genomes for performing analyses such as PICRUSt (REFS). Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling is problematic. As one illustration of the effects that incorrect labels would have on an analysis, we asked whether the duplicate sequences had the same taxonomies. Among the `r sum(dups$V2==2)` reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies. Furthermore, among those `r sum(dups$V2 > 2)` sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies. Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies. Among the sequences within the database, there was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies. When we analyzed the `r n_actual_map` sequences that mapped to the reference sequences using a non-heuristic approach, we observed that `r n_map_more_than_one_expanded` of the sequences mapped to more than one reference sequence. Of these sequences, `r n_map_one_uniqued` (`r 100*n_map_one_uniqued/n_map_more_than_one_expanded`%) mapped to references that were identical over the V4 region and `r n_map_more_than_one_uniqued` (`r 100*n_map_more_than_one_uniqued/n_map_more_than_one_expanded`%) mapped equally well to two or more references that were not identical over the V4 region. Among the combined `r n_map_more_than_one_expanded` sequences that mapped to multiple reference sequences, the taxonomy of the multiple reference sequences conflicted for `r n_mismatch_taxonomy` (`r 100*n_mismatch_taxonomy/n_map_more_than_one_expanded`%). Together, these results demonstrate some of the considerable problems with the reference-based clustering of sequences.

@@

Our analysis of those methods that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its specificity. When we replaced USEARCH with VSEARCH, the clustering quality improved. Although there may be parameters in USEARCH that can be tuned to improve the heuristic, these parameters are likely dataset dependent. Based on the data presented in this study, its availability as an open source, and free program, VSEARCH should replace USEARCH in these clustering methods. Furthermore, although not tested in our study, VSEARCH can be parallelized leading to potentially significant improvements in speed. Although USEARCH and VSEARCH do not utilize aligned sequences, it is important to note that a sequence curation pipeline including denoising, alignment, trimming to a consistent region of the 16S rRNA gene, and chimera checking are critical to making proper inferences (REF).

We have assessed the ability of reference-based clustering methods to capture the actual distance between the sequences in a dataset in parallel with *de novo* methods. Several studies have lauded both the open and closed-reference approaches for generating reproducible clusterings (REFS), yet we have shown that both reference-based approaches did a poor job of representing the distance between the sequences compared to the *de novo* approaches. Although the OTU assignments are reproducible and stable across a range of library sizes, the reference-based OTU assignments are a poor representation of the data. We also observed that the assignments were not actually reproducible when the order of the reference sequences was randomize. When USEARCH was used, the actual number of sequences that mapped to the reference changed depended on the order of the reference. Perhaps most alarming was that the default order of the database provided the worst MCC of any of the randomizations we attempted. Even when we used VSEARCH to perform closed-reference clustering and were able to obtain a consistent clusterings, we observed that the labels on the OTUs differed between randomizations. Because the OTU labels are frequently used to identify representative sequences for those OTUs, variation in labels, often representing different taxonomic groups, will have a detrimental effect on the interpretation of downstream analyses.

>>>>

---
title: "***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

\begin{center}
\vspace{25mm}
Patrick D. Schloss${^\dagger}$ and Sarah L. Westcott

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage
\linenumbers

## Abstract

@@

The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples. To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins. These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@Schloss2011]. Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g. @Huttenhower2012], soil [e.g. @Shade2013], aquatic ecosystems [e.g. @Gilbert2011], and numerous other environments. Within the field of microbial ecology, a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%. One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project. However, with the dissemination of clustering methods within software such as mothur [@Schloss2009], QIIME [@Caporaso2010], and other tools [@Edgar2010; @Sun2009; @Mah2014; @Edgar2013; @Cai2011], it is important to understand how different clustering methods implement this conventional OTU threshold. Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs. Broadly speaking, three approaches have been developed to assign sequences to OTUs.

The first approach has been referred to as phylotyping [@Schloss2011] or closed reference clustering [@NavasMolina2013]. This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@Schloss2011]. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.

The second approach has been referred to as distance-based [@Schloss2011] or *de novo* clustering [@NavasMolina2013]. In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures then these problems can be overcome [@Kozich2013]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@Edgar2010, @Sun2009, @Mah2014]. One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [@He2015; @Hamady2009]. Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011]. For this reason, *de novo* clustering has been preferred across the field.

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013]. Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions employed by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014]. These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering. Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @Chen2013; @Edgar2013]. This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists. Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010].  Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014; @Rideout2014]. This is problematic because it does not solve the fundamental question of which method is most correct. We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011]. Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient [MCC; @Matthews1975]. This is an objective approach to assessing the quality of the OTU assignments.

A recent analysis by He and colleagues @He2015 attempted to characterize the three general clustering approaches by focusing on what they called stability. They defined stability as the ability of an method to provide the same clustering on a subset of the data as was found in the full dataset. Their concept of stability did not account for the accuracy of the OTU assignments and instead focused on the precision of the assignments. A method may be very precise, but low in accuracy. In the current analysis, we assessed the accuracy and precision of the various clustering methods. Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of accuracy. In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.

@@

***Summary and replication of He study.*** We obtained the Canadian soil dataset from Roesch et al. @Roesch2007 and processed the sequences as described by He and colleagues. Using these data, we reconsidered three of the more critical analyses performed in the He study.

@@

Third, the authors attempted to describe the effects of the OTU assignment instability on comparisons of communities. They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset when only using the unstable OTUs [@Anderson2001]. Although they were able to detect significant p-values, they appeared to be of marginal biological significance. Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different. The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL). Regardless of the statistical or biological significance of these results, the analysis does not make sense since, by definition, representing communities based on their unstable OTUs would yield differences. Furthermore, the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds. To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets. The justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable. It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels. Finally, the authors built off of this analysis to count the number of OTUs that were differentially represented between the subsetted and full datasets by each method. This analysis assumes that the OTUs generated using the full dataset were correct, which was an unsubstantiated assumption since the authors did not assess the quality of the OTU assignments. Because this analysis was so poorly designed, we did not seek to reproduce it.

***OTU assignment methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences. To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth. This enabled us to calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C). The SL and CL methods were the most sensitive; however, the accuracy of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods. Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D). Third, with the possible exception of the CL method, the MCC values for each of the only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same method, they produce OTU assignments that are equally good. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the the full dataset is not optimally assigned to OTUs. Our analysis indicates that the best method for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL method.

@@

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods. First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permit the sequencing of millions of sequences for a study. In addition, because the original Phred quality scores and flowgram data are not available, it was not possible for us to adequately remove sequencing errors [@Schloss2011Reducing]. The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods. Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene [@Schloss2010]. With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012; @Kozich2013]. The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% [@Kozich2013]. Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points. First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the He dataset (Figure 3AB). With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`. Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the Canadian soil dataset for all of the methods except the CL and SL methods. Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD). This result suggests that the optimal method is likely to be database-dependent. Finally, as was observed with the Canadian soil dataset, there was little variation in MCC values observed among the 30 randomizations. Therefore, although the methods have a stochastic component, the OTU assignments do not vary meaningfully between runs. The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the assignments are highly reproducible. Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods are consistently more robust than either of the reference-based methods.

***Evalution of an open-source alternative to USEARCH.*** For some datasets the AGC and DGC methods appear to perform as well or better than the hierarchical clustering methods. As originally described in the He study, the AGC and DGC methods utilized the USEARCH program [@Edgar2010], which is not available as open source code and is only available for free to academic users as a 32-bit program. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to generate candidate alignments whereas VSEARCH generates the actual global alignments. The VSEARCH developers claim that this difference enhances the sensitivity of VSEARCH relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC methods, as implemented by the two programs, yielded OTU assignments of similar quality. In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the methods (Figure 4). When we compared the two implementations of the AGC and DGC methods, the OTUs generated by the VSEARCH-verion of the methods were as stable or more stable than the USEARCH-version when using 60% of the datasets. In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable. These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC methods.

@@

Beyond comparing the abundance of specific OTUs across samples, the reference database is used in the open and closed-reference methods to generate OTU labels that are used in several downstream applications. It is commonly used to extract information from a reference phylogenetic tree to carrying out UniFrac-based analyses [@Hamady2009] and to identify reference genomes for performing analyses such as PICRUSt [@Langille2013]. Because these downstream applications depend on the correct and unique labeling of the OTUs, the lack of stability of the labeling is problematic. As one illustration of the effects that incorrect labels would have on an analysis, we asked whether the duplicate sequences had the same taxonomies. Among the `r sum(dups$V2==2)` reference sequences that had one duplicate, `r sum(dups$V2==2 & dups$V3==2)` had discordant taxonomies. Furthermore, among those `r sum(dups$V2 > 2)` sequences with two or more duplicates, `r sum(dups$V2>2 & dups$V3>=2)` had discordant taxonomies. Two sequences mapped to `r max_taxa[1,2]-1` and `r max_taxa[2,2]-1` duplicate sequences and both contained `r max_taxa[1,3]` different taxonomies. Among the sequences within the database, there was also a sequence had `r max(dups[,2])-1` duplicates and contained `r dups[which.max(dups[,2]),3]` different taxonomies. When we analyzed the `r n_actual_map` sequences that mapped to the reference sequences using a non-heuristic approach, we observed that `r n_map_more_than_one_expanded` of the sequences mapped to more than one reference sequence. Of these sequences, `r n_map_one_uniqued` (`r 100*n_map_one_uniqued/n_map_more_than_one_expanded`%) mapped to references that were identical over the V4 region and `r n_map_more_than_one_uniqued` (`r 100*n_map_more_than_one_uniqued/n_map_more_than_one_expanded`%) mapped equally well to two or more references that were not identical over the V4 region. Among the combined `r n_map_more_than_one_expanded` sequences that mapped to multiple reference sequences, the taxonomy of the multiple reference sequences conflicted for `r n_mismatch_taxonomy` (`r 100*n_mismatch_taxonomy/n_map_more_than_one_expanded`%). Together, these results demonstrate some of the considerable problems with the reference-based clustering of sequences.

@@

Our analysis of those methods that implemented USEARCH as a method for clustering sequences revealed that its heuristic limited its specificity. When we replaced USEARCH with VSEARCH, the clustering quality improved. Although there may be parameters in USEARCH that can be tuned to improve the heuristic, these parameters are likely dataset dependent. Based on the data presented in this study, its availability as an open source, and free program, VSEARCH should replace USEARCH in these clustering methods. Furthermore, although not tested in our study, VSEARCH can be parallelized leading to potentially significant improvements in speed. Although USEARCH and VSEARCH do not utilize aligned sequences, it is important to note that a sequence curation pipeline including denoising, alignment, trimming to a consistent region of the 16S rRNA gene, and chimera checking are critical to making proper inferences [@Schloss2011Reducing; @Kozich2013; @Schloss2012Secondary].
